---
output:
  html_document: default
  pdf_document: default
---


Prosper Loan Analysis using R
=======================================================
#### By Xia Chen, 2018-11-10, Toronto

# 1.Introduction
> Prosper is America's first marketplace peer-to-peer lending platform, with over $12 billion in funded loans. This data set contains 113,937 loans with 81 variables on each loan, including loan amount, borrower rate (or interest rate), current loan  status, borrower income, borrower employment status, borrower credit score, etc. The explaination of all the variables in the data can be found [here](https://docs.google.com/spreadsheets/d/1gDyi_L4UvIrLTEC6Wri5nbaMmkGmLQBk-Yx3z0XDEtI/edit?usp=sharing)

> In this project, an Exploratory Data Analysis (EDA) for the Prosper loan data set was conducted, in order to dig out some inspiring and valuable insights of the loan. Then Machine learning algorithms was applied to predict the default risk. 

#### The outline of this project is:
  
>1.  Introduction
1.1 Background
1.2 Install R packages
1.3 Data loanding and exploring
2.  Data Cleaning
2.1 Missing values
2.2 Drop columns and rows
2.3 Filling null values with column mean
3.  Exploratory Data Analysis
3.1 Univariate Plots Section
3.2 Bivariate Plots Section
3.3 Multivariate Plots Section
4. Feature Engineering
4.1 Data pre-processing
4.2 Feature selection
4.3 Model validation and selection
4.4 Default prediction
5. Conclusion and summary

## 1.1 Install R packages

> In this analysis, I used 10 packages for data manipulation, data visulization and correlation analysis. 
> The name are below:
ggplot2, date, dplyr, gridExtra, scales, choroplethrMaps, chroplethr, GGally, knitr.

```{r echo=FALSE,message=FALSE, warning=FALSE}
library(knitr)
library(ggplot2)
library(dplyr)
library(date)
library(GGally)
library(gridExtra)
library(scales)
library(choroplethrMaps, quietly =TRUE)
library(choroplethr, quietly = TRUE)
library(rmarkdown)
library(corrplot)
library(caret)
library(fastDummies)
library(randomForest)
```


## 1.2 Load data

```{r echo=FALSE,message=FALSE, warning=FALSE}

loan<-read.csv('C:/Users/admin/Documents/R/prosperLoanData.csv')
cat("There are ", dim(loan)[1], "observations of", dim(loan)[2], "variables.")
```


> 1.The Prosper loan data contains 113,937 loans that were issued through 2005-2014, with 81 variables for each loan. 
2. The target is the loan status which could help investors estimate the default risk of borrowers.
3. The interesting features that might affect the loan status could from numerical variables, including borrower rate, monthly income,loan amount, credit score, term of the loan, total prospers loans.
The fetures could also from and categorical variables, including employment status, loan purpose, home ownership, borrower state.

# 2 Data processing and cleaning 

## 2.1 Duplicates

```{r echo=FALSE,message=FALSE, warning=FALSE}
cat("There are ", sum(duplicated(loan)), "dupplicates")
```

##2.2 Missing values

```{r echo=FALSE,message=FALSE, warning=FALSE}
colSums(is.na(loan))
```

- There are 9 columns with over 90,000 missing values; some columns are either irrelevant or beyound the scope of this analysis. So they need to be removed. 

##2.3 Drop some irrelevant columns

```{r echo=FALSE,,message=FALSE, warning=FALSE}
# drop irrelevant columns
loan<-subset(loan, select=-c(ListingKey,ListingNumber,LoanKey,LoanNumber, MemberKey))

# drop columns with missing values over 90000
loan<-subset(loan, select=-c(TotalProsperLoans,TotalProsperPaymentsBilled,OnTimeProsperPayments,ProsperPaymentsLessThanOneMonthLate,ProsperPaymentsOneMonthPlusLate,ProsperPrincipalBorrowed,ProsperPrincipalOutstanding,ScorexChangeAtTimeOfListing,LoanFirstDefaultedCycleNumber))

# drop columns describe the status of a loan at present or the history of the loan. these are beyond the scope of this analysis.
loan<-subset(loan, select=-c(LoanCurrentDaysDelinquent,  LoanMonthsSinceOrigination, LP_CustomerPayments, LP_CustomerPrincipalPayments, LP_InterestandFees, LP_ServiceFees, LP_CollectionFees, LP_GrossPrincipalLoss, LP_NetPrincipalLoss, LP_NonPrincipalRecoverypayments))

# drop some time series column
loan<-subset(loan, select=-c(ClosedDate, DateCreditPulled,FirstRecordedCreditLine, LoanOriginationDate, LoanOriginationQuarter ))

# drop columns that I don't think is relevent to loan status. For the prosper rating and prosper score, they only applicable for loans after July 2009, so I will drop them as well. Income range  is less useful than montly income because the ranges are very broad. 
loan<-subset(loan, select=-c(CreditGrade, BorrowerAPR, LenderYield, EstimatedEffectiveYield, EstimatedLoss, EstimatedReturn,ProsperRating..numeric., ProsperRating..Alpha., ProsperScore,Occupation, CurrentlyInGroup, GroupKey,CurrentCreditLines,TotalCreditLinespast7years, IncomeRange,IncomeVerifiable, PercentFunded,InvestmentFromFriendsCount,InvestmentFromFriendsAmount ))


```

## 2.4 Drop some rows with missing values

```{r echo=FALSE,message=FALSE, warning=FALSE}

loan <- loan[!is.na(loan$TradesOpenedLast6Months ), ]
loan <- loan[!is.na(loan$BankcardUtilization ), ]
loan <- loan[!is.na(loan$ EmploymentStatusDuration ), ]
loan <- loan[!is.na(loan$  CurrentDelinquencies ), ]
colSums(is.na(loan))
```

- After dropping the columns, several other columns have over 7544 missing values. So I removed them. 

## 2.5 Filling  missing values with mean

```{r echo=FALSE,message=FALSE, warning=FALSE}

loan$DebtToIncomeRatio<- ifelse(is.na(loan$DebtToIncomeRatio), mean(loan$DebtToIncomeRatio, na.rm=TRUE), loan$DebtToIncomeRatio)

```

- Now the data is pretty clean, but there is still 8404 missing values for debt to income ratio, and this is a very important variable, so I will fill the missing values with mean values. 

##2.6 Tidy column names and variables 

```{r echo=FALSE,message=FALSE, warning=FALSE}
# change column name
colnames(loan)[colnames(loan)=="ListingCategory..numeric."]<- "ListingCategory"
colnames(loan)[colnames(loan)=="TradesNeverDelinquent..percentage."]<- "TradesNeverDelinquent"
# change date type 
loan$LoanCreationDate<- as.Date(loan$ListingCreationDate, formate="%m/%d/%Y")
loan<-subset(loan, select=-c(ListingCreationDate))
colSums(is.na(loan))
```

- After cleaning up the missing values, there are some column names and date type nee to tidy up.

##2.7 Check data type

```{r echo=FALSE,message=FALSE, warning=FALSE}
cat("There are ", dim(loan)[1], "observations of", dim(loan)[2], "variables.")

```

- After data cleaning step, now There are  106290 observations of 33 variables.

#3. Exploratory Data Analysis

## 3.1 Univariate Plots Section

> In this section, I want to start exploring the data by visulization
The fetures that I am interested to investigate are below:

### 3.1.1 Loan Status

> In risk management one important job is to build a predictive model to  predict whether the loan will be default or not.  So the loan status wil be the target variable. However, there are 12 loan status. But we need to label them as default or non defaul loans. 'delinquent', 'default' and 'charge off' are terms helping investor to estimate whether or not they can expect to collect on the outstanding debt at all. According to the definition of delinquencies from [badcredit.org](https://www.badcredit.org/delinquencies-defaults-charge-offs-whats-difference/), an account will be considered severely delinquent if no payment has been received 60 days beyond the due date.Those delinquent accounts may have higher risk of default.

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(LoanStatus),data=loan)+
  geom_bar(stat="count")+
  ggtitle("Loan Status")+
  theme(plot.title = element_text(hjust = 0.5))+
  coord_flip()
```

### 3.1.2 Define delinquent borrower

> So I created a new variable 'DelinquentBorrowers' using '0' to represent none delinquent borrowers and '1' to represent delinquent borroweres, including loans that are noted default,  charged off, and past due over 60 days. According to   [LendingClub](https://help.lendingclub.com/hc/en-us/articles/216127747-What-is-the-difference-between-a-loan-that-is-in-default-and-a-loan-that-has-been-charged-off-), In general, a note goes into default status  when it is 121 or more days past due. When a note is in default status,  Charge Off occurs no later than 150 days past due (i.e. No later than 30  days after the Default status is reached) when there is no reasonable  expectation of sufficient payment to prevent the charge off.


```{r echo=FALSE, message=FALSE, warning=FALSE}
# define delinquent borrowers
loan$DelinquentBorrowers <- ifelse(
loan$LoanStatus == "Defaulted" |
loan$LoanStatus == "Chargedoff" |
loan$LoanStatus == "Past Due (61-90 days)" |
loan$LoanStatus == "Past Due (91-120 days)" |
loan$LoanStatus == "Past Due (>120 days)", 
 1, 0)
ggplot(aes(DelinquentBorrowers, fill=DelinquentBorrowers),data=loan)+
  geom_bar(stat="count")+
  ggtitle('Delinquent borrowers')+
  theme(plot.title = element_text(hjust = 0.5))
```

- There is a high proportion of borrowers who are currently delinquent at least once. But the table summarized that only 13% of borrowers have severe delinquent records. 

```{r echo=FALSE, message=FALSE, warning=FALSE}
# calculate percentage of deliquencies
percent_delinquent<-loan%>%
                 group_by(DelinquentBorrowers)%>%
                 summarise(counts=n())%>%
                 mutate(Percentage=counts/sum(counts)*100)
percent_delinquent 
```

### 3.1.3 Histogram plot

```{r echo=FALSE, message=FALSE, warning=FALSE}
# create histogram of loan amount
loan_amount_median <- median(loan$LoanOriginalAmount)
loan_amount_hist <- ggplot(loan, aes(LoanOriginalAmount)) + 
             geom_vline(xintercept=loan_amount_median)+
             geom_histogram(binwidth=750, fill="#E69F00", colour="black") +               xlim(2500,25000)+
             xlab("Amount borrowed (USD)") +
             ylab("Number of loans") + 
             ggtitle("Histogram of loan amounts")+
             theme_set(theme_gray(base_size = 10))+
             annotate("text", x = loan_amount_median + 4800, y = 17000,  
             label = paste("Median =",loan_amount_median, "USD"), size=3)
 
# creat histogram of monthly income 
monthly_income_median<- round(median(loan$StatedMonthlyIncome))
monthly_income_hist <- ggplot(loan, aes(StatedMonthlyIncome)) +
             geom_histogram(binwidth=750, fill="#E69F00", colour="black") +               xlim(0,20000)+
             xlab("Monthly income (USD)") +
             ylab("Number of loans") + 
             ggtitle("Histogram of monthly income")+
             theme_set(theme_gray(base_size = 10))+
             geom_vline(xintercept=monthly_income_median)+
             annotate("text", x = monthly_income_median + 4800, y = 17000,  
             label = paste("Median =",monthly_income_median, "USD"), size=3)

# creat histogram of interest rate
loan$BorrowerRate<-loan$BorrowerRate*100
borrower_rate_median<-median(loan$BorrowerRate)
borrower_rate_hist <- ggplot(loan, aes(BorrowerRate)) +
             geom_histogram( fill="#E69F00", colour="black") +  
             xlim(0,40)+
             ylim(0,10000)+
             xlab("Borrower rate (%)") +
             ylab("Number of loans") + 
             ggtitle("Histogram of interest rate")+
             theme_set(theme_gray(base_size = 10))+
             geom_vline(xintercept=borrower_rate_median)+
             annotate("text", x = borrower_rate_median + 5, y = 10000,
             label = paste("Median =",borrower_rate_median, "%"), size=3)

 
# creat histogram of credit score
credit_score_median<- median(loan$CreditScoreRangeLower)
credit_score_hist <- ggplot(loan, aes(CreditScoreRangeLower)) +
             geom_histogram(binwidth=20,fill="#E69F00", colour="black") + 
             xlim(400,900)+
             xlab("Credit score") +
             ylab("Number of loans") + 
             ggtitle("Histogram of credit score")+
             theme_set(theme_gray(base_size = 10))+
             geom_vline(xintercept=credit_score_median)+
             annotate("text", x = credit_score_median + 50, y = 17000, 
             label = paste("Median =",credit_score_median), size=3)

grid.arrange(loan_amount_hist,monthly_income_hist, borrower_rate_hist,  credit_score_hist, ncol=2)

```

- There are four important features that I believe will tell the borrower's 
ability to pay back the loan on time. They are the loan amount, monthly income, interest rate, and credit score. This histogram plots above give me an idea that most loans are small loans,so it make sense the interest rate might be higher. Indeed, 18.4% interest rate is quite high. Most borrowers' are meddle class with monthly income around 5000 dollars, and their credit score is good, because the median value is 700.  

### 3.1.4 Numeric univariate plots

```{r echo=FALSE, message=FALSE, warning=FALSE}
credit_utilization<-ggplot(aes(BankcardUtilization),data=loan)+
  geom_histogram()+
  xlim(0,1)+
  ggtitle("Borrowers's bank credit utilization")

available_credit<-ggplot(aes(AvailableBankcardCredit),data=loan)+
  geom_histogram()+
  xlim(0,50000)+
  ggtitle("Borrowers's available bank credit")

current_delinquencies<-ggplot(aes(CurrentDelinquencies),data=loan)+
  geom_histogram()+
  xlim(0,10)+
  ylim(0,15000)+
  ggtitle("CurrentDelinquencies")

loan_terms<-ggplot(aes(Term),data=loan)+
  geom_histogram()+
  xlim(0,60)+
  ggtitle(" Loan terms")


home_owner<-ggplot(aes(IsBorrowerHomeowner),data=loan)+
  geom_histogram(stat="count")+
  ggtitle('Homeowner of the loan borrowers')+
  theme(plot.title = element_text(hjust = 0.5))

debt_to_income<-ggplot(aes(DebtToIncomeRatio),data=loan)+
  geom_histogram()+
  ggtitle("Debt to income ratio")+
  xlim(0, 1)+
  theme(plot.title = element_text(hjust = 0.5))

 grid.arrange(credit_utilization, available_credit,current_delinquencies,loan_terms,home_owner,debt_to_income, ncol=2)
```

### 3.1.5 Categorical univariables

>The loan categories are give as numerical variables, in order to investigate the correlation of loan purpose with other features, the 'ListingCategory' column was coverted into categorical variable.

```{r echo=FALSE, message=FALSE, warning=FALSE}

#creat new column of loan categories
loan$NewListingCat<-cut(loan$ListingCategory, breaks=c(0,1,2,3,4,5,6,7,8,9,10,11,12,13,14,15,16,17,18,19,20,21),
                        labels=c( 'Not Available', 'Debt Consolidation', 'Home Improvement',  'Business', 'Personal Loan', 'Student Use',  'Auto', 'Other', 'Baby&Adoption', 'Boat', 'Cosmetic Procedure', 'Engagement Ring', 'Green Loans', 'Household Expenses', 'Large Purchases', 'Medical/Dental', 'Motorcycle', 'RV', 'Taxes','Vacation','Wedding Loans'), right=FALSE)
#plot loan purpose
loan_purpose<-ggplot(aes(NewListingCat),data=loan)+
ggtitle('Loan Purpose')+
  coord_flip()+
  geom_bar(stat="count")+
  theme(plot.title = element_text(hjust = 0.5))
#plot employment status
employment_status<-ggplot(aes(EmploymentStatus),data=loan)+
  geom_bar(stat="count")+
  coord_flip()+
  ggtitle('Employment status')+
  theme(plot.title = element_text(hjust = 0.5))

grid.arrange(loan_purpose, employment_status, ncol=2)
```

- From the histogram plots for categorical variables, we get to know that there is high volumn of loans are for the purpose of debt consolidation. So later I need to investigate their creditline and other banck credit utilization informatin.


##3.2 Bivariate Plots Section

### 3.2.1 Correlation heatmap

```{r echo=FALSE, message=FALSE, warning=FALSE, correlation_plot2, fig.width=15, fig.height=8}

loan_new<-loan%>%select( Term, LoanStatus, BorrowerRate,           ListingCategory,EmploymentStatusDuration, IsBorrowerHomeowner,OpenCreditLines,CurrentDelinquencies, CreditScoreRangeLower,BankcardUtilization,TotalInquiries, DelinquenciesLast7Years, AvailableBankcardCredit,DebtToIncomeRatio,StatedMonthlyIncome,LoanOriginalAmount)

ggcorr(loan_new, label = TRUE, label_round = 2, label_size =3.5, size =5, hjust = .85) +
  ggtitle("Correlation Heatmap") +
  theme(plot.title = element_text(hjust = 0.5))

```

- 14 variables are selected to investigate the correlation between those variables. Borrower rate has some correlation with employment status duration, bankcard utilization, bank card utilization, and monthly income. 

### 3.2.2 Employment Status vs.Monthly income

```{r echo=FALSE, message=FALSE, warning=FALSE}

 loan$StatedMonthlyIncome.bucket = cut(loan$StatedMonthlyIncome,
                            c(0, 3200, 4667, 5608 ,  6825, 1750000))

ggplot(aes(x = StatedMonthlyIncome.bucket,
           y = LoanOriginalAmount), 
           data=subset(loan, (StatedMonthlyIncome < 11000) & 
                      (EmploymentStatus != ""))) + 
   geom_boxplot(aes(fill = EmploymentStatus) )   +
      theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

-The monthly income is related to employment status, I cut the montly income into 6 levels and created a new variable MonthlyIncome.bucket. Interestingly, not employed does not mean low monthly income, employed and self-employed people also have very high monthly income. 

### 3.2.3  Geographical distribution of loans 

```{r echo=FALSE, message=FALSE, warning=FALSE}
data(state.regions)
loan<-merge(loan, state.regions, by.x="BorrowerState", by.y="abb")

deliquencies_by_state<-loan %>%
  group_by(region)%>%
  filter(region != "NA") %>% 
  na.omit()%>%
  summarise(value=sum(DelinquentBorrowers))
state_choropleth(deliquencies_by_state, title="Deliquencies by state", num_colors=9)



```

- From a geographical perspective California, Texas, New York, Florida and Illinois have the largest dollar amounts and volumes of loans. The delinquent loans is geographically consistent with the high volume and large amount of the loan.  But the average loan amount is higher in states Alaska and South Carolina, I think this might be because of the lower level of pupulation in these two states. 

## 3.3 Multivariate Plots Section

```{r echo=FALSE,message=FALSE, warning=FALSE}
#2.4.1 Homeownership vs. Credit score
p11<-ggplot(aes(x=IsBorrowerHomeowner, y=CreditScoreRangeLower,
      color = factor(DelinquentBorrowers)), data=loan) +
  geom_point(alpha = 0.1, position = "jitter") +
  scale_color_discrete(labels = c("Good loan", "Delinquent loan")) +
  theme(legend.title=element_blank())+
  ggtitle('Homeownership vs. Credit score')+
  theme(plot.title = element_text(hjust = 0.5))


# 2.4.2 Loan purpose vs. Credit score
p12<-ggplot(aes(x=NewListingCat, y=CreditScoreRangeLower,
       color = factor(DelinquentBorrowers)), data=loan) +
  ylim(500,900)+
  geom_point(alpha = 0.1, position = "jitter") +
  scale_color_discrete(labels = c("Good loan", "Delinquent loan")) +
  theme(legend.title=element_blank())+
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1))+
  ggtitle(' Loan purpose vs. Credit score')+
  theme(plot.title = element_text(hjust = 0.5))

grid.arrange(p11,p12, ncol=2)

```

```{r echo=FALSE,message=FALSE, warning=FALSE}

# 2.4.3 Credit score vs. Borrower rate
p13<-ggplot(aes(x=CreditScoreRangeLower, y=BorrowerRate,
      color = factor(DelinquentBorrowers)), data=loan) +
  scale_x_continuous(limits = c(500, 900)) +
  scale_y_continuous(limits = c(0, 40)) +
  geom_point(alpha = 0.1, position = "jitter") +
  scale_color_discrete(labels = c("Good loan", "Delinquent loan")) +
  theme(legend.title=element_blank())+
  ggtitle('Credit score vs. Borrower rate')+
  theme(plot.title = element_text(hjust = 0.5))


#2.4.4 Credit score vs. Employment status
p14<-ggplot(aes( x=EmploymentStatus, y=CreditScoreRangeLower,
      color = factor(DelinquentBorrowers)), data=loan) +
  scale_y_continuous(limits = c(500,900)) +
  geom_point(alpha = 0.1, position = "jitter") +
  scale_color_discrete(labels = c("Good loan", "Delinquent loan")) +
  theme(legend.title=element_blank())+
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1))+
  ggtitle(' Employment status vs. Credit score')+
   theme(plot.title = element_text(hjust = 0.5))

grid.arrange(p13,p14, ncol=2)
```

```{r echo=FALSE,message=FALSE, warning=FALSE}

#2.4.5 Credit score vs. Employment status duration
p15<-ggplot(aes( x=EmploymentStatusDuration, y=CreditScoreRangeLower,
      color = factor(DelinquentBorrowers)), data=loan) +
  scale_y_continuous(limits = c(500,900)) +
  geom_point(alpha = 0.1, position = "jitter") +
  scale_color_discrete(labels = c("Good loan", "Delinquent loan")) +
  theme(legend.title=element_blank())+
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1))+
  ggtitle(' Employment status duration vs. Credit score')+
   theme(plot.title = element_text(hjust = 0.5))

# 2.4.6 Credit score vs. Loan amount
p16<-ggplot(aes(x=CreditScoreRangeLower, y=LoanOriginalAmount,
      color = factor(DelinquentBorrowers)), data=loan) +
  scale_x_continuous(limits = c(500,900)) +
  scale_y_continuous(limits = c(0, 50000)) +
  geom_point(alpha = 0.1, position = "jitter") +
  scale_color_discrete(labels = c("Good loan", "Delinquent loan")) +
   theme(legend.title=element_blank())+
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1))+
   ggtitle(' Credit score vs. Loan amount')+
   theme(plot.title = element_text(hjust = 0.5))
grid.arrange(p15,p16, ncol=2)
```

```{r echo=FALSE,message=FALSE, warning=FALSE}

# 2.4.7 Credit score vs. monthly income
p17<-ggplot(aes(x=CreditScoreRangeLower, y=StatedMonthlyIncome,
      color = factor(DelinquentBorrowers)), data=loan) +
  scale_x_continuous(limits = c(500,900)) +
  scale_y_continuous(limits = c(0, 50000)) +
  geom_point(alpha = 0.1, position = "jitter") +
  scale_color_discrete(labels = c("Good loan", "Delinquent loan")) +
   theme(legend.title=element_blank())+
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1))+
   ggtitle(' Credit score vs. Monthly income')+
   theme(plot.title = element_text(hjust = 0.5))


# 2.4.8 Credit score vs. Debt-to-income ratio
p18<-ggplot(aes(x=CreditScoreRangeLower, y=DebtToIncomeRatio,
      color = factor(DelinquentBorrowers)), data=loan) +
  scale_x_continuous(limits = c(500,900)) +
  scale_y_continuous(limits = c(0, 1)) +
  geom_point(alpha = 0.1, position = "jitter") +
  scale_color_discrete(labels = c("Good loan", "Delinquent loan")) +
  theme(legend.title=element_blank())+
  ggtitle(' Credit score vs. Debt-to-income ratio')+
 theme(plot.title = element_text(hjust = 0.5))
grid.arrange(p17,p18, ncol=2)
```

- 1. Borrower rate is related to prosper score, higher borrower rate leads to lower Prosper score. Delinquent borrowers might due to higher borrower rate.

- 2. Deliquent borrowers are small loan borrowers. Also, deliquent borrowers tend to have higher Debt-to-income ratio. 

## 3.4  Reflection

>1. The prospser loans dataset contains over 100k observations with 81 variables spanning across 10 years.The first step before conducting any data analysis is to understand the variables, terminology and general domain knowledge of financial peer-to-peer lending.Second, it is very important to determine which variables to analyze and stick to those variables without drifting too far off.Also there are so many missing values and none specific observations, cleaning is needed for looking into the relationship between some variables.

>2. For loan data, I believe that default risk analysis is a key component to help investors to decide if they could collect the full loan.So any features that could lead to default are worth investigating, In my analysis, I found that borrower rate from investor's aspect and Prosper score from borrower's aspect are strongly relatted to deliquencies. This brought me to investigate more features that could possibly affect borrower rate and prosper score, including monthly income, borrower region, loan purpose, employment status, credit card utilization, available credit, loan amount,home ownership, and debt to income ratio. 

## 3.5 Limitations

> The Exploratory Data Analysis strategy is a good way to find some insights of the date through interesting visualizations. However, EDA works better for data set with limited variables. For the Prosper Loan data with 81 variables, it is really time consuming to find out the most intesting features and investigate the correlations between them.  In the future,an useful method to investigate this data could be building predictive models using machine learning. 

# 4 Default prediction using machine learning

## 4.1 Data-preprocessing
### 4.1.1 Create dummy variables

```{r echo=FALSE,message=FALSE, warning=FALSE}
#Creat a new data frame for  prediction

loan_predict<-subset(loan, select=-c(LoanStatus, BorrowerState,NewListingCat,                               StatedMonthlyIncome.bucket, region, 
                   fips.numeric,fips.character,LoanCreationDate)) 


#Variables EmplymentStatus and ISBorrowerHomeowner are factors, I will convert them into nemerical variables.

## convert boolean values into 0 and 1. If the borrower is home owner,covert 'true' into 1, else convert it into 0. 
loan_predict$Homeownership <- ifelse(loan_predict$IsBorrowerHomeowner == "True",  1, 0)

## convert employment status into dummy variables
loan_predict$EmploymentStatus<-as.character(loan_predict$EmploymentStatus)

loan_predict$EmploymentStatus[loan_predict$EmploymentStatus == "Not available"] <- "Other" 

loan_predict<-fastDummies::dummy_cols(loan_predict, select_columns='EmploymentStatus')

## rename dummy variables

colnames(loan_predict)[colnames(loan_predict)=="EmploymentStatus_Self-employed"]<- "EmploymentStatus_Selfemployed"

colnames(loan_predict)[colnames(loan_predict)=="EmploymentStatus_Full-time"]<- "employed_fulltime"

colnames(loan_predict)[colnames(loan_predict)=="EmploymentStatus_Part-time"]<- "EmploymentStatus_Parttime"

colnames(loan_predict)[colnames(loan_predict)=="EmploymentStatus_Not employed"]<- "EmploymentStatus_Notemployed"


#Change data types of dummy variables
loan_predict[33:39]<-lapply(loan_predict[33:39], as.numeric)

## combine enmplymentstatus 'employed=employed+full_time
loan_predict$Employment_Status_Employed<-loan_predict$EmploymentStatus_Employed  + loan_predict$employed_fulltime +loan_predict$EmploymentStatus_Selfemployed

# Drop some columns after converting 
loan_predict<-subset(loan_predict, select=-c(IsBorrowerHomeowner,EmploymentStatus, EmploymentStatus_Employed, employed_fulltime,EmploymentStatus_Selfemployed))
write.csv(loan_predict, file='loan_predict.csv')

```

### 4.1.2 Remove Highly Correlated Variables
```{r echo=FALSE,message=FALSE, warning=FALSE,fig.width=30, fig.height=16}
# Check the correlation of all variables in loan_predict
ggcorr(loan_predict, label = TRUE, label_round = 2, label_size =3.5, size =5, hjust = .85) +
  ggtitle("Correlation Heatmap") +
  theme(plot.title = element_text(hjust = 0.5))

# remove highly correlated variables

loan_predict<-subset(loan_predict, select=-c(EmploymentStatus_Other,CreditScoreRangeUpper, TotalInquiries, OpenRevolvingAccounts,MonthlyLoanPayment,OpenRevolvingMonthlyPayment))
loan_predict$DelinquentBorrowers<-as.factor(loan_predict$DelinquentBorrowers)

```

- Since the problem is a classification problem, I choose logistic regression, random forest and Adaboosting.

## 4.2 Model validation and selection

### 4.2.1 Model comparison and evaluation by running K-Fold cross validation 

- Since this project is a binnary classification problem, three algorithms were selected to compare: Logistic Regression, Naive Bayes, and Random Forest. The accuracy of those models were compared by running K-fold cross validation.
- From the plot above, random forest give the best accuracy, which is 0.87.

###4.2.2 Parameter tuning
- In this section, two very important parameters mtry and ntree are tuned.
- The error rate is lowestat 12.39% s when  mtry =7.
- OOB is bottoming out at approximately 1000 trees. So ntree=1000 would be good fit 

###4.2.3 Final Random Forest Classification

- With the tunned parameters, the accuracy increased to 100%. 

# 5 Conclusion
> 1. The process of analyzing Prosper loan data is a good experience in data cleanning, visulizaion, and feature engineering. 
> 2. The advantages of random forest modeling include telling people about the importance of each variables.
> 4. Some other features not included it the dataset could be also very important, such as the borrower's age, and gender. 
> 3. This project could go further by carefully selecting the features. Future work could be predicting the loss and profit, predicting the interest rate to minimize the default, etc. 

# 6 References

- 1. https://www.earthdatascience.org/courses/earth-analytics/time-series-data/summarize-time-series-by-month-in-r/
- 2. https://stackoverflow.com/questions/33221425/how-do-i-group-my-date-variable-into-month-year-in-r
- 3. https://www.r-bloggers.com/plot-weekly-or-monthly-totals-in-r/
- 4. https://github.com/cran/choroplethr/blob/master/R/choropleth.R
- 5. https://stackoverflow.com/questions/41376522/cregion-value-in-colnamesuser-df-are-not-all-true
- 6. https://stats.stackexchange.com/questions/13465/how-to-deal-with-an-erro
- 7. https://stackoverflow.com/questions/5824173/replace-a-value-in-a-data-frame-based-on-a-conditional-if-statement
- 8. https://cran.r-project.org/web/packages/fastDummies/vignettes/making-dummy-variables.html
- 9. https://campus.datacamp.com/courses/machine-learning-with-tree-based-models-in-r/boosted-trees?ex=12
- 10. https://machinelearningmastery.com/compare-the-performance-of-machine-learning-algorithms-in-r/
- 11. https://www.r-bloggers.com/how-to-implement-random-forests-in-r/
- 12. https://stackoverflow.com/questions/32849479/run-randomforest-with-many-mtry-values
- 13. https://stackoverflow.com/questions/35655701/is-there-a-quicker-way-of-running-gridsearchcv